Acceleration of plague outbreaks in the second pandemic

### 2020-07-04 ms_text.tex
minor typos

%%\begin{abstract}
{\bfseries
\noindent
Historical records reveal the temporal patterns of a sequence of plague epidemics in London, England, from the 14th to 17th centuries.  Analysis of these records shows for the first time that later epidemics spread significantly faster (``accelerated'').  Between the Black Death of 1348 and the later epidemics that culminated with the Great Plague of 1665, we estimate that the epidemic growth rate increased \foldval.
%%[\rdiffest $\times$, 95\% CI: \rdifflwr $\times$ to \rdiffupr $\times$].  
%%The cause of this change is unknown; despite rapid recent advances in paleogenomics, it is hard to confirm or reject the possibility of functional evolutionary changes in circulating strains of the plague bacterium \ypestis over this time period.
Currently available data do not provide enough information to infer the mode of plague transmission in any given epidemic; nevertheless, order-of-magnitude estimates of epidemic parameters suggest that the observed slow growth rates in the 14th century are inconsistent with direct (pneumonic) transmission.  
We discuss the potential roles of demographic and ecological factors such as climate change or human or rat population density in driving the observed acceleration.
}  
%%\end{abstract}

@@

We can use the distribution of the generation interval (the times elapsed between the moment a host is infected and the times at which they infect others) to compute $\rzero$ from $r$ \cite{WallLips07}. If we only know the \emph{mean} generation interval $\gentime$, we can still approximate $\rzero \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\rzero$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.

@@

The growth rates for the 14th century epidemics are lower than those for the 16th-17th centuries (\cref{F:combdata}); they may also be more variable than in the later period, although some of the appearance of variability is due to larger uncertainties (presumably due to the relatively poor sampling in the early epoch). The mixed model quantifies the dramatic increase in growth rate (``acceleration'') of epidemics between epochs.
The growth rates increased \foldval in the late \emph{vs.}\ early epochs: $r$ increased \rdiffest $\times$ [95\% CI $(\rdifflwr\times, \rdiffupr\times)$] (see \Methodslink). The permutation test results support the statistical significance of this difference; the observed order of outbreaks yields the most extreme difference in growth rates out of all \nperm\ possible orderings, giving a two-tailed $p$ value of $\permfrac=\permpval$.

@@

Why the later plague epidemics were faster than the earlier ones is not clear.  Nevertheless, to provide some context, we consider below a number of mechanisms which \emph{could} have contributed to increases in epidemic growth rates.  We then consider what our estimated epidemic growth rates suggest about the mode of transmission during the various London plague epidemics.

@@

Now suppose that the late-epoch plagues were pneumonic with a generation interval similar to that of Modern pneumonic plague.  The implied reproduction number (\cref{F:combdata}) would be $\rzerop=\rzerolateest$ [CI (\rzerolatelwr, \rzerolateupr)],
which yields a maximum attack rate of \Zlateest\%  [CI (\Zlatelwr\%, \Zlateupr\%)], compared to 
the estimated $\approx 20\%$ of the population
who died in each of the late epoch epidemics \cite[p.\,4]{Cumm+16}.
Given that the CFP for (Modern) pneumonic plague is nearly 100\%
\cite{KoolWein05,WHOplaguePage,WHOplagueFactSheet,CDCplagueEmergencyFAQ},
reconciling the predicted attack rates with observed mortality would require strong effects of heterogeneity; a large fraction of individuals resistant to plague due to prior non-fatal plague infections; a large proportion of asymptomatic or sub-clinical infections; or a decrease in transmission rate as the plague spread (perhaps due to behavioural avoidance of transmission).
The effects of heterogeneity and behavioural avoidance may indeed have been stronger in the late epoch, as a result of concentration of the wealthy in central London and tendency for them to flee London during plague epidemics \cite[p.\,4]{Cumm+16}. While we still consider direct (pneumonic) transmission in  late-epoch London plague epidemics unlikely because of the strong effects that would be needed to reduce the expected mortality rates to the observed levels, we cannot rule it out.

>>>>

%%\begin{abstract}
{\bfseries
\noindent
Historical records reveal the temporal patterns of a sequence of plague epidemics in London, England, from the 14th to 17th centuries.  Analysis of these records shows for the first time that later epidemics spread significantly faster (``accelerated'').  Between the Black Death of 1348 and the later epidemics that culminated with the Great Plague of 1665, we estimate that the epidemic growth rate increased \foldval.
%%[\rdiffest $\times$, 95\% CI: \rdifflwr $\times$ to \rdiffupr $\times$].  
%%The cause of this change is unknown; despite rapid recent advances in paleogenomics, it is hard to confirm or reject the possibility of functional evolutionary changes in circulating strains of the plague bacterium \ypestis over this time period.
Currently available data do not provide enough information to infer the mode of plague transmission in any given epidemic; nevertheless, order-of-magnitude estimates of epidemic parameters suggest that the observed slow growth rates in the 14th century are inconsistent with direct (pneumonic) transmission.  
We discuss the potential roles of demographic and ecological factors, such as climate change or human or rat population density, in driving the observed acceleration.
}  
%%\end{abstract}

@@

We can use the distribution of the generation interval (the times elapsed between the moment a host is infected and the times at which they infect others) to compute $\rzero$ from $r$ \cite{WallLips07}. If we know only the \emph{mean} generation interval $\gentime$, we can still approximate $\rzero \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\rzero$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.

@@

The growth rates for the 14th century epidemics are lower than those for the 16th-17th centuries (\cref{F:combdata}); they may also be more variable than in the later period, although some of the appearance of variability is due to larger uncertainties (presumably due to the relatively poor sampling in the early epoch). The mixed model quantifies the dramatic increase in growth rate (``acceleration'') of epidemics between epochs.
The growth rates increased \foldval in the late \emph{vs.}\ early epochs: $r$ increased \rdiffest $\times$ [95\% CI $(\rdifflwr\times, \rdiffupr\times)$] (see \Methodslink). The permutation test results support the statistical significance of this difference; the observed order of outbreaks yields the most extreme difference in growth rates out of all \nperm\ possible orderings, giving a two-tailed $p$-value of $\permfrac=\permpval$.

@@

Why the later plague epidemics were faster than the earlier ones is not clear.  Nevertheless, to provide some context, we consider below a number of mechanisms that \emph{could} have contributed to increases in epidemic growth rates.  We then consider what our estimated epidemic growth rates suggest about the mode of transmission during the various London plague epidemics.

@@

Now suppose that the late-epoch plagues were pneumonic with a generation interval similar to that of Modern pneumonic plague.  The implied reproduction number (\cref{F:combdata}) would be $\rzerop=\rzerolateest$ [CI (\rzerolatelwr, \rzerolateupr)],
which yields a maximum attack rate of \Zlateest\%  [CI (\Zlatelwr\%, \Zlateupr\%)], compared to 
the estimated $\approx 20\%$ of the population
who died in each of the late epoch epidemics \cite[p.\,4]{Cumm+16}.
Given that the CFP for (Modern) pneumonic plague is nearly 100\%
\cite{KoolWein05,WHOplaguePage,WHOplagueFactSheet,CDCplagueEmergencyFAQ},
reconciling the predicted attack rates with observed mortality would require strong effects of heterogeneity; a large fraction of individuals resistant to plague due to prior non-fatal plague infections; a large proportion of asymptomatic or sub-clinical infections; or a decrease in transmission rate as the plague spread (perhaps due to behavioural avoidance of transmission).
The effects of heterogeneity and behavioural avoidance may indeed have been stronger in the late epoch, as a result of concentration of the wealthy in central London and a tendency for them to flee London during plague epidemics \cite[p.\,4]{Cumm+16}. While we still consider direct (pneumonic) transmission in  late-epoch London plague epidemics unlikely because of the strong effects that would be needed to reduce the expected mortality rates to the observed levels, we cannot rule it out.
